Vertical and oblique transmission with fluctuating transmission

### 2018-06-10 ms/ms_sunnyvale.tex
finished migration proof

For the simple case of symmetric migration $m_1=m_2=m<1/2$ after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

We have the following results:
\begin{itemize}
\item With only oblique transmission ($\rho_1=\rho_2=0$), the only equilibria are the neutral equilibria $(x^*,x^*)$ for any $0 \le x^* \le 1$.
\item With only vertical transmission ($\rho_1=\rho_2=1$), the fixation equilibria  $(0,0)$, $(1,1)$ are unstable and there exists a protected polymorphism
\begin{equation}
\begin{aligned} \label{eq:migration_model_unconditional_symmetric_polymorphism}
x_1^* &= \frac{w-1-m(w+1) + \sqrt{\Delta}}{2(w-1)}, \\
x_2^* &= 1-x_1^*,
\end{aligned}
\end{equation}
where $\Delta = m^2(w+1)^2+(1-2m)(w-1)^2$.
\item With only vertical transmission in one deme ($\rho_1=1$) and a combination of both vertical and oblique transmission in the other deme ($\rho_2=\rho$), fixation of $B$ is unstable, and fixation of $A$ is stable if and only if the vertical transmission rate in deme 2 is below $\hat \rho$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_rho}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)},
\end{equation}
or if the migration rate is above $\hat m$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m}
m > \hat m = \frac{\rho (w-1)}{\rho (w-1) +1-\rho}.
\end{equation}
\end{itemize}

@@

For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $f(\lambda) = a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
First, the discriminant $\Delta = b^2-4ac$ is positive for any real $m$, because, when writing it as a quadratic polynomial in $m$, $\Delta(m)=\frac{1}{w^{2}} \big(m^{2} (w^2+1)^2 - 2 m (w^2-1)^2 + (w^2-1)^2\big)$, this polynomial is positive at $m=0$ and $m=1/2$ and has a negative discriminant $-16w^2(w^2-1)^2$.
Therefore, the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two positive roots. \\
Second, as $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial $f(\lambda)$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w + 1-\rho + \rho/w), \quad
c &=  (1-2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}

First, the discriminant is 
$$
\Delta = (1 - m)^2 (w + 1-\rho + \rho/w)^2 - 4(1-2m)(w(1-\rho)+\rho),
$$
which is positive for $m=0$ and $m=1/2$:
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w + \rho) (w-1)}{w}\Big)^2 \\
\Delta(m=1/2) = \Big(\frac{w^2 + w(1-\rho) + \rho}{2 w}\Big)^2.
\end{aligned}\end{equation}
In addition, if we write $\Delta$ as a polynomial in $m$, its discriminant is negative -- $- 16 (w + \rho)^2 (w - 1)^2 \big(w (1 - \rho) + \rho \big) / w^2$, and therefore the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1-m)(w \rho + 1 -\rho + 1/w), \quad
c &= (1-2m) \big((1-\rho)/w + \rho\big),
\end{aligned}
\end{equation}
%[MWF: Are the roots real?]
Again, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur in $(0,1)$ and fixation is stable. %

For any $\rho$, $a+b+c>0$ if and only if
\begin{equation}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)},
\end{equation}
and then $(1,1)$ is stable if and only if $2a+b>0$.
We have already seen that when $\rho=0$ we have $2a+b>0$.
When $\rho = \hat \rho$, we have $2a+b=\frac{m(w+1)+w-1}{w(1-m)(w-1)}>0$.
Since $2a+b$ is linear in $\rho$, we find that $2a+b>0$ for $\hat \rho \ge \rho \ge 0$ and therefore \eqref{eq:migration_model_unconditional_symmetric_condition_rho} guarantees the stable fixation of $A$.

The values of $a+b+c$ at $m=0$ and $m=1$ are $-\frac{\rho(w-1)^2}{w}<0$ and $\frac{(1-\rho)(w-1)}{w}>0$, respectively, and $a+b+c$ is linear in $m$, so that we have $a+b+c>0$ when 
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m2}
m > \hat m = \frac{\rho (w-1)}{\rho (w-1) +1-\rho}.
\end{equation}
In addition, $2a+b$ is linear in $m$ and equal to 2 when $m=1$; it is equal to zero when $m=(w-1)(w\rho-1)/(w(w-1)\rho+w+1) < \hat m$, and therefore (\ref{eq:migration_model_unconditional_symmetric_condition_m}, also \ref{eq:migration_model_unconditional_symmetric_condition_m2})
is sufficient for stability of fixation of $A$.

@@

% Facebook rural internet 
% National service - Israel, Singapore
% Immigrtion - draught
% Opening schools, universities
% sending students abroad, bringing experts from abroad
% foreign labor
% refugees
% hunter-gatherers/chimps increase habitat in years with less food production
% religion - dark ages etc.

% Acknowledgements
%{\small
%\section*{Acknowledgements}
%
%This work was supported in part by 
%the Stanford Center for Computational, Evolutionary and Human Genomics, 
%and the Morrison Institute for Population and Resources Studies, Stanford University.
%}

>>>>

For the case of symmetric migration $m_1=m_2=m$ ($0\le m \le1/2$) occurring after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

We have the following results:
\begin{itemize}
\item With only oblique transmission ($\rho_1=\rho_2=0$), the only equilibria are the neutral equilibria $(x^*,x^*)$ for any $0 \le x^* \le 1$.
\item With only vertical transmission ($\rho_1=\rho_2=1$), the fixation equilibria  $(0,0)$, $(1,1)$ are unstable and there exists a protected polymorphism
\begin{equation}
\begin{aligned} \label{eq:migration_model_unconditional_symmetric_polymorphism}
x_1^* &= \frac{w-1-m(w+1) + \sqrt{\Delta}}{2(w-1)}, \\
x_2^* &= 1-x_1^*,
\end{aligned}
\end{equation}
where $\Delta = m^2(w+1)^2+(1-2m)(w-1)^2$.
\item With only vertical transmission in one deme ($\rho_1=1$) and a combination of both vertical and oblique transmission in the other deme ($\rho_2=\rho$), fixation of $B$ is unstable, and fixation of $A$ is stable if and only if the vertical transmission rate in deme 2 is below $\hat \rho$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_rho}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)} < 1,
\end{equation}
or if the migration rate is above $\hat m$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m}
m > \hat m = \frac{\rho w - 1}{\rho w + 1}.
\end{equation}
\end{itemize}

@@

%For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $f(\lambda) = a \lambda^2 + b \lambda + c$ with
%\begin{equation}
%\begin{aligned}
%a &= 1, \quad
%b &= -\frac{(1-m)(w^2+1)}{w}, \quad
%c &=  1-2m.
%\end{aligned}
%\end{equation}
%First, the discriminant $\Delta = b^2-4ac$ is positive for any real $m$, because, when writing it as a quadratic polynomial in $m$, $\Delta(m)=\frac{1}{w^{2}} \big(m^{2} (w^2+1)^2 - 2 m (w^2-1)^2 + (w^2-1)^2\big)$, this polynomial is positive at $m=0$ and $m=1/2$ and has a negative discriminant $\big(-16w^2(w^2-1)^2\big)$.
%Therefore, the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two positive roots. \\
%Second, as $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, with $0 \le \rho \le 1$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial of the Jacobian of~\eqref{eq:migration_model_unconditional_symmetric}, $f_0(x)=ax^2+bx+c$, with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w + 1 - \rho + \rho/w), \quad
c &=  (1 - 2m)(\rho + w(1-\rho)).
\end{aligned}
\end{equation}

First, the discriminant $\Delta = b^2-4ac$ for $m=0$ and $m=1/2$ is
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w + \rho) (w-1)}{w}\Big)^2, \text{and} \\
\Delta(m=1/2) = \Big(\frac{w^2 + w(1-\rho) + \rho}{2 w}\Big)^2,
\end{aligned}\end{equation}
which are both positive.
Writing $\Delta$ as a polynomial in $m$, $g(m)$, the discriminant of $g(m)$ is 
$$
\big(- 16 (w + \rho)^2 (w - 1)^2 \big(w (1 - \rho) + \rho \big) / w^2\big),
$$
which is negative, and therefore the discriminant $\Delta$ of $f_0(x)$ is positive for any $0 \le m \le 1/2$ and $f_0(x)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ so $f_0$ is positive with a negative derivative at $x=0$ and has a positive derivative at infinity.
Therefore stability of $(0,0)$ can be determined by the sign of $f_0(1)=a+b+c$.
For $\rho=1$ we have $a+b+c=-\frac{1-m}{w}(w-1)^2<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any $0 \le \rho \le 1$, so $f(x)$ has a real root greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial $f_1(x)=ax^2+bx+c$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w\rho + 1 - \rho + 1/w), \quad
c &=  (1 - 2m)(\rho + (1-\rho)/w).
\end{aligned}
\end{equation}

First, the discriminant $\Delta = b^2-4ac$ for $m=0$ and $m=1/2$ is
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w \rho + 1) (w-1)}{w}\Big)^2, \text{and} \\
\Delta(m=1/2) = \Big(\frac{w^2 \rho + w(1-\rho) + 1}{2 w}\Big)^2,
\end{aligned}\end{equation}
which are both positive. 
Writing the discriminant $\Delta$ of $f_1(x)$ as a polynomial in $m$, $g(m)$, the discriminant of $g(m)$ is
$$
\big(- 16 (w\rho + 1)^2 (w - 1)^2 \big(\rho w + 1 - \rho \big) / w^3\big),
$$
which is negative, and therefore $\Delta$ is positive for any $0 \le m \le 1/2$ and $f_1(x)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ so $f_1$ is positive with a negative derivative at $x=0$ and has a positive derivative at infinity.
Therefore, fixation of $A$ is stable iff both $f_1(1)=a+b+c$ and $f_1'(1)=2a+b$ are positive.
$f_1(1)>0$ occurs if $\rho < \frac{m}{m+(1-m)(w-1)}$.
$f_1'(1)>0$ occurs if $\rho < \frac{m(w+1)+w-1}{(1-m)w(w-1)}$, which is larger than $\frac{m}{m+(1-m)(w-1)}$ for any $w>1$ and $0 \le m \le 1/2$ (the difference is a quadratic in $w$ with a negative discriminant $2m-1$ and negative value $-1/2$ at $w=2$, $m=0$). 
Therefore, fixation of $A$ is stable iff $\rho < \frac{m}{m+(1-m)(w-1)}$ \eqref{eq:migration_model_unconditional_symmetric_condition_rho}, or, by rearranging this inequality, we can also get an expression for $m$ \eqref{eq:migration_model_unconditional_symmetric_condition_m}.

@@

% Facebook rural internet 
% National service - Israel, Singapore
% Boarding schools
% Immigrtion - draught
% Opening schools, universities
% sending students abroad, bringing experts from abroad
% foreign labor
% refugees
% hunter-gatherers/chimps increase habitat in years with less food production
% religion - dark ages etc.

% Acknowledgements
{\small
\section*{Acknowledgements}

This work was supported in part by 
the Stanford Center for Computational, Evolutionary and Human Genomics, 
and the Morrison Institute for Population and Resources Studies, Stanford University.
}
